---
title: "Functional Programming with R"
author: Julen Astigarraga y Verónica Cruz-Alonso
date: today
date-format: "DD/MM/YYYY"
toc: true
format:
  html: default
    # css: styles.css
  gfm: default
editor: visual
editor_options: 
  chunk_output_type: console
number-sections: false
---

## 2.2.1 Exercise

1.  With the penguins dataset, create an object with individuals of the Adelie species and arrange the `data.frame` according to flipper length.

2.  Create an object from the previous one where you select the island and the variables related to the bill.

3.  Create an object from the one created in step 1 where you select everything except the species.

```{r 2.2.1}

library(palmerpenguins)
library(tidyverse)

# 1
adelie <- penguins |> 
  filter(species == "Adelie") |> 
  arrange(flipper_length_mm)

# 2
adelie_subset <- adelie |> 
  select(island, contains("bill"))

# 3
adelie2 <- adelie |> 
  select(-species)

```

## 2.2.2 Exercise

1.  With the penguins dataset, count the number of cases on each island and calculate the mean flipper length for each island.

2.  Calculate the ratio between the weight in kg and the flipper length for each individual.

```{r 2.2.2}

# 1
penguins |> 
  group_by(island) |> 
  summarise(
    n = n(),
    mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE)
  )

# 2
penguins |> 
  mutate(mass_by_length = (body_mass_g / 1000) / flipper_length_mm) |> 
  select(mass_by_length)

```

## 5.0.1 Exercise

Create your first function that fulfills the functionality of dividing any value by 100.

```{r 5.0.1}

divide_by_100 <- function(x) {
  x / 100
}

divide_by_100(30)

```

## 5.0.2 Exercise

Create a function to standardize (i.e., subtract the mean and divide by the standard deviation) the numerical variables of the penguins dataset.

```{r 5.0.2}

# R base
standardise <- function(x, my.na.rm) {
  (x - mean(x, na.rm = my.na.rm)) / sd(x, na.rm = my.na.rm)
}

standardise(penguins$bill_length_mm, my.na.rm = T)

# Tidyverse
standardise_tidy <- function(x, my.na.rm) {
  ({{ x }} - mean({{ x }}, na.rm = my.na.rm)) / sd({{ x }}, na.rm = my.na.rm)
} 

penguins |> 
  mutate(st_bill_length = standardise_tidy(x = bill_length_mm, my.na.rm = T)) |> 
  select(bill_length_mm, st_bill_length)

```

## 5.2.1 Exercise

How would you generalize the function `explore_penguins()` so you can use any data frame as input?

```{r 5.2.1}

penguins_num <- penguins |> 
  select(species, sex, where(is.numeric))

explore_penguins <-
  function(var) {
    ggplot(penguins_num, aes(x = species, y = .data[[var]], color = sex)) +
      geom_point(position = position_jitterdodge(), alpha = 0.3) +
      geom_boxplot(alpha = 0.5) +
      scale_color_manual(values = c("turquoise", "goldenrod1")) +
      theme_light() +
      ylab(var)
  }

explore_penguins("body_mass_g")

explore_df <- function(df, x_var, y_var, color) {
  ggplot(data = df, aes(x = .data[[x_var]], y = .data[[y_var]], color = .data[[color]])) + 
    geom_point(position = position_jitterdodge(), alpha = 0.3) +
    geom_boxplot(alpha = 0.5) + 
    scale_color_manual(values = rainbow(n = length(unique(df[[color]])))) +
    theme_light() + 
    ylab(y_var)
}

explore_df(df = iris, # iris es un dataset cargado en RStudio por defecto
           y_var = "Petal.Length",
           x_var = "Species",
           color = "Species")

```

<details>

<summary>Session Info</summary>

```{r session_info}
Sys.time()
sessionInfo()
```

</details>
